---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Sys.setlocale("LC_TIME", "C"); # Setlocale is important if you have non-English locale
```

## Loading and preprocessing the data
1. Load the data (i.e. ```read.csv()```)
2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r activity}
unzip(zipfile = "activity.zip")
activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
1. Make a histogram of the total number of steps taken each day
```{r steps.total}
steps.total <- tapply(activity$steps, 
                     activity$date, 
                     FUN = sum, 
                     na.rm = TRUE)

hist(steps.total, breaks=40, 
     main="Total number of steps taken each day", 
     xlab="Days", ylab="Number of steps")
```

2. Calculate and report the **mean** and **median** total number of steps taken per day
```{r mean}
mean(steps.total, na.rm = TRUE)
```
```{r median}
median(steps.total, na.rm = TRUE)
```



## What is the average daily activity pattern?
1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r averages}
steps.averages <- aggregate(x = list(steps = activity$steps), 
                            by = list(interval = activity$interval), 
                            FUN = mean, 
                            na.rm = TRUE)

plot(steps.averages, type='l', main='Average Daily Activity Pattern',
     ylab='Steps', xlab='Intervals')
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r maxnumberofsteps}
steps.averages$interval[which.max(steps.averages$steps)]
```

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with ``NAs``)
```{r missing}
missingFunc <- function(values) sum(is.na(values))
sapply(activity, missingFunc)
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

<i>As it can be seen from the output above we have missing values only in `steps` field. I decided to change `NA` values by the average number of steps for such field.</i>
```{r replacedNAs}
#Get NA's indexes and duplicate table
missingRows <- is.na(activity$steps)
replacedNAs <- activity

#Create key-value dictionary with interval-steps pairs
keyvalue <- as.list(steps.averages[[2]])
names(keyvalue) <- as.list(steps.averages[[1]])

#Replace NA values by the values from the dictionary
intervalsOfMissing <- as.character(activity$interval[missingRows])
replacedSteps <- sapply(intervalsOfMissing, 
                        FUN = function(x) {keyvalue[[x]]}, 
                        USE.NAMES = FALSE)

#Update the table
replacedNAs$steps[missingRows] <- replacedSteps
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r replacedNAsPlot}
steps.total <- tapply(replacedNAs$steps, replacedNAs$date, FUN = sum)
hist(steps.total, breaks=40, 
     main="Total number of steps taken each day", 
     xlab="Days", ylab="Number of steps")
```
```{r meanReplacedNAs}
mean(steps.total, na.rm = TRUE)
```
```{r medianReplacedNAs}
median(steps.total, na.rm = TRUE)
```
<i>The mean and median have been changed and the distribution lost its skewness. Now the distribution has a look more similar to normal distribution, because **median=mean**. The reason why previous distribution had another values is because the total number of steps taken in NA fields are set to 0s by default. So we had skewness towards the left side.</i>

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r weekdays}
determine.weekend <- function(date) {
  day <- weekdays(date)
  if (day %in% c("Saturday", "Sunday")) return("Weekend")
  else return("Weekday")
}

#Find all weekends in the table
replacedNAs$date <- as.Date(replacedNAs$date)
replacedNAs$days <- sapply(replacedNAs$date, determine.weekend)
```

2. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r showbydays}
library(lattice)
steps.averages <- aggregate(steps ~ interval + days, data = replacedNAs, mean)
xyplot(steps~interval|days, type='l', data = steps.averages, layout=c(1,2), xlab = "Intervals", ylab = "Steps", col = 'black')
```
